---
title: "bidag_comparison"
output: html_document
date: '2022-07-13'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Comparing DAGMC to BiDAG

## Setup

```{r setup, message = FALSE}
library(dagmc)
library(bnlearn)
library(dplyr)
library(ggplot2)
```

Get data. Just using categorical learning.test data at this point. Comparisons
with Gaussian data seem aren't consistent, which I think is due to BiDAG using
the updated score in Kuipers, Moffa, & Heckerman (2014).

```{r cars}
data <- bnlearn::learning.test[1:100, ]
type <- 'bde'
# true_graph <- bnlearn::empty.graph(names(data))
# arc_set <- matrix(c('A', 'F',
#                     'A', 'C',
#                     'B', 'C',
#                     'B', 'D',
#                     'G', 'F',
#                     'D', 'F',
#                     'E', 'F'),
#                   ncol = 2, byrow = TRUE,
#                   dimnames = list(NULL, c('from', 'to')))
```


# ```{r, error = TRUE}
# true_graph <- bnlearn::empty.graph(names(data))
# bnlearn::arcs(true_graph) <- arc_set
# true_adj <- true_graph %>%
#  bnlearn::as.igraph() %>%
#  igraph::as_adjacency_matrix() %>%
#  as.matrix
# bnlearn::graphviz.plot(bnlearn::cpdag(true_graph))
# ```

## Posterior sampling


### Using DAGMC

Using 

```{r, error = TRUE}
system.time(
  scorer <- CreateScorer(scorer = BNLearnScorer,
                       data = data, type = type,
                       max_parents = Inf,
                       blacklist = NULL,
                       whitelist = NULL,
                       cache = TRUE)
)
```

```{r, error = TRUE}
set.seed(1)
n_results <- 10000

init_dag <- matrix(0L, 
                   nrow = ncol(data), ncol = ncol(data), 
                   dimnames = list(names(data), names(data)))
init_partition <- GetPartitionedNodesFromAdjacencyMatrix(init_dag)

system.time(
  chain <- SampleChain(n_results, init_partition,
                      transition = PartitionMCMC(),
                      scorer = scorer)
)
```

Sample DAGs from partitions.
```{r, error = TRUE}
system.time( chain <- SampleChainDAGs(chain, scorer) )
```


### Using BiDAG

```{r, error = TRUE}
bd_scorer <- BiDAG::scoreparameters(scoretype = 'bdecat',
                                    data = data, 
                                    bdecatpar = list(chi = 1.0, edgepf = 1.0))

startspace <- matrix(1L, 
                     nrow = ncol(data), ncol = ncol(data), 
                     dimnames = list(names(data), names(data)))

system.time(
  bd_res <- BiDAG::partitionMCMC(bd_scorer,
                                 iterations = n_results,
                                 moveprobs = c(0.33, 0.0, 0.34, 0.33, 0.0),
                                 stepsave = 1,
                                 startspace = startspace,
                                 verbose = TRUE)
)
```

## Compare the two


Check for convergence.

```{r, error = TRUE, fig.dim = c(6.0, 4.0)}
PlotScoreTrace(chain, attribute = 'log_dag_score', ylab = 'log(DAG score)', type = 'l', col = 'blue')
lines(bd_res$trace)
```

Did they agree on probabilities for each DAG?

```{r, error = TRUE}
p_dagmc <- chain$dag %>%
  sapply(rlang::hash) %>%
  table %>%
  "/"(n_results) %>%
  data.frame %>%
  tibble %>%
  dplyr::rename(hash_dag = 1)

bd_dags <- BiDAG::getTrace(bd_res, which = 1) %>%
  lapply(function(x) as(x, 'matrix')) %>%
  sapply(function(x) rlang::hash(x)) %>%
  table %>%
  "/"(sum(.)) %>%
  data.frame %>%
  tibble %>%
  dplyr::rename(hash_dag = 1)

dplyr::full_join(p_dagmc, bd_dags, by = 'hash_dag')
```

Do they have the same DAG probabilities?


Check for convergence after burnin.

```{r, error = TRUE}
PlotScoreTrace(eq_chains, ylab = 'log(partition score)', type = 'l')
```

Sample DAGs from partitions and plot the trace.
```{r, error = TRUE}
system.time(
  eq_chains <- SampleChainDAGs(eq_chains, scorer)
)
PlotScoreTrace(eq_chains, attribute = 'log_dag_score', ylab = 'log(DAG score)', type = 'l')
```

# Posterior sampling using BiDAG



